\chapter{System Implementation}

\section{Reviewing the Design Solution}

The design solution was reviewed before implementation to confirm that it still aligns with the project objectives and remains practical given real-world constraints (performance, reliability, and maintainability). This review focused on verifying that each architectural decision supports the core learning workflow and that the system can evolve without costly redesign.

\textbf{Alignment with project objectives}

\begin{itemize}
    \item \textbf{Personalized learning paths:} The ontology combined with the progress model supports adaptive learning paths, mastery state transitions, and multi-path branching. This aligns directly with the goal of personalized roadmaps.
    \item \textbf{Adaptive assessment:} The separation of quiz generation (ai-service) from quiz delivery (learning-service) ensures assessments can adapt while maintaining consistent delivery and scoring rules.
    \item \textbf{AI-grounded content and adaptation:} The two-layer content architecture (offline expert curation + runtime ontology-grounded AI generation) aligns with the goal of providing high-quality, hallucination-free explanations. The adaptation subsystem (resource\_swap, prerequisite\_review, instructor\_escalation) provides targeted remediation based on failure depth.
\end{itemize}

\textbf{Architecture validity check}

\begin{itemize}
    \item \textbf{Service separation remains justified:} The gateway isolates authentication and RBAC; the learning-service owns domain rules; the ai-service manages external AI dependencies. This avoids cross-cutting security and reliability risks.
    \item \textbf{Scalability assumptions hold:} AI workloads are isolated and can scale horizontally without affecting database-heavy learning flows.
    \item \textbf{Data integrity is preserved:} Prisma schema and constraints enforce the DAG structure, progress tracking, and versioning rules required by the ontology model (see \texttt{backend/services/learning-service/prisma/schema.prisma}).
\end{itemize}

\textbf{Design trade-offs revisited}

\begin{itemize}
    \item \textbf{Microservices vs.\ monolith:} A monolith would reduce operational overhead but would couple AI latency and auth concerns to core learning logic. The current split avoids cascading failures and allows targeted scaling.
    \item \textbf{Relational database choice:} PostgreSQL provides strong consistency and relational constraints needed for prerequisite validation and mastery tracking. This is essential for correctness and auditability.
    \item \textbf{AI dependency management:} The design anticipates AI provider instability and includes circuit breakers and caching. This trade-off increases implementation complexity but improves availability.
\end{itemize}

\textbf{Implementation readiness}

\begin{itemize}
    \item \textbf{Deployment model continuity:} The Docker Compose and Nginx routing match the deployment diagram, ensuring local development mirrors production patterns.
    \item \textbf{Tooling fit:} TypeScript, Express, and Prisma align with the need for fast iteration, schema safety, and consistent API definitions in the gateway and learning service. Python 3.11 with FastAPI and Pydantic was selected for the ai-service due to Python's first-class AI/ML ecosystem and FastAPI's native async generator support for SSE streaming.
\end{itemize}

\textbf{Design refinements confirmed during implementation}

\begin{itemize}
    \item \textbf{Input validation tightened:} Validation and sanitization at the gateway were strengthened to enforce contracts and reduce security risks.
    \item \textbf{Caching depth increased:} AI response caching was expanded to reduce latency and improve throughput under load.
    \item \textbf{Error handling standardized:} Common error objects and response formats were enforced across services for consistent client behavior.
    \item \textbf{SSE streaming added:} Real-time token-by-token streaming was introduced for AI explanations and the AI Instructor chatbot, with Nginx \texttt{proxy\_buffering off} and gzip disabled on stream routes.
    \item \textbf{Certificate migration applied out-of-band:} The certificates table was applied via raw SQL rather than \texttt{prisma migrate dev} to avoid a \texttt{UserRole} enum drift that would have deleted production instructor accounts.
    \item \textbf{Gamification wired as fire-and-forget:} XP awards and badge checks are dispatched asynchronously from the gatekeeper to avoid adding latency to the quiz submission critical path.
\end{itemize}

Overall, the design remains valid and aligned with the project objectives. The refinements above did not change the architecture but improved its operational robustness and implementation clarity.

\section{Deciding on the Development Tools}

The development toolchain was selected to balance rapid iteration, reliability, and long-term maintainability. The choices reflect the system's need for strong typing, modular services, a consistent API layer, and cross-platform client delivery.

\subsection{Programming Languages}

\begin{itemize}
    \item \textbf{TypeScript:} Used across the api-gateway, learning-service, and web frontend to provide static typing, safer refactoring, and shared conventions.
    \item \textbf{Python 3.11:} Used exclusively for the ai-service, which orchestrates LLM provider calls, SSE streaming, prompt engineering, Redis caching, and circuit-breaker logic. Python was chosen for its first-class AI/ML ecosystem (google-generativeai, httpx, Pydantic) and its ergonomic support for async generators, which map naturally onto Server-Sent Events token streams.
    \item \textbf{Dart:} Chosen for the mobile client to leverage Flutter's cross-platform UI toolkit and strong developer productivity.
    \item \textbf{SQL (via Prisma):} The relational model requires precise constraints and indexing; Prisma generates SQL migrations while keeping schema changes declarative.
\end{itemize}

\subsection{Frameworks and Core Libraries}

\begin{itemize}
    \item \textbf{Node.js + Express (api-gateway, learning-service):} Express provides lightweight routing and middleware, enabling clear separation of controllers, services, and validation layers.
    \item \textbf{FastAPI (ai-service):} FastAPI is used as the async web framework for the Python AI service. Its native support for \texttt{StreamingResponse} and async generators makes SSE token streaming idiomatic; Pydantic model validation enforces request/response contracts at the framework level without additional schema libraries.
    \item \textbf{Pydantic (ai-service):} Request bodies, AI provider responses, and internal service contracts are validated using Pydantic models, providing Python-native type-safe data parsing equivalent to Joi on the TypeScript side.
    \item \textbf{Uvicorn (ai-service):} ASGI server that runs the FastAPI application; supports HTTP/1.1 with keep-alive connections required for SSE.
    \item \textbf{google-generativeai (ai-service):} Official Google Python SDK for Gemini API, supporting both synchronous generation and streaming via \texttt{generate\_content\_stream}.
    \item \textbf{httpx (ai-service):} Async HTTP client used for Phi-4 and Ollama provider calls, with timeout and connection pool configuration.
    \item \textbf{redis-py (ai-service):} Async Redis client for AI response caching and circuit-breaker state storage.
    \item \textbf{ORM (Prisma):} Used by learning-service for schema consistency, migration tracking, and type-safe queries.
    \item \textbf{Validation (Joi):} Input schemas enforce contract integrity in the TypeScript services.
    \item \textbf{Logging (Pino / Python logging):} Structured logging supports monitoring and audit trails across services.
\end{itemize}

\subsection{AI and External Integration Tools}

\begin{itemize}
    \item \textbf{LLM Providers:} Three-tier fallback chain --- Phi-4-Multimodal (primary), Qwen2.5-3B via Ollama (secondary, local CPU), and Gemini 2.5 Flash API (tertiary, last resort). Circuit breakers per provider prevent cascading failures.
    \item \textbf{Redis caching:} AI responses are cached with per-type TTLs (explanations 24h, quizzes 7d) to avoid repeat generation costs.
    \item \textbf{Server-Sent Events (SSE):} Used for token-by-token streaming of AI explanations and AI Instructor chat responses.
    \item \textbf{SERP API:} Used for supplementary resource discovery scoped to ontology node learning outcomes.
\end{itemize}

\subsection{Frontend and Mobile Tooling}

\begin{itemize}
    \item \textbf{Frontend:} React 18 with Vite provides fast HMR builds and a modern component model. React Router manages navigation. Key libraries: Zustand (client state), TanStack Query (server state and caching), Tailwind CSS (utility-first styling), React Flow (DAG roadmap visualisation), Recharts (analytics charts), Zod + React Hook Form (form validation), html-to-image + jsPDF (certificate download).
    \item \textbf{Mobile:} Flutter provides a consistent UI across Android and iOS, with centralised state and theme control; GraphView renders the learning DAG on mobile.
\end{itemize}

\subsection{Testing and Quality Tooling}

\begin{itemize}
    \item \textbf{Jest:} Unit and integration tests for backend services.
    \item \textbf{Playwright:} End-to-end testing for full user journeys.
    \item \textbf{ESLint/Prettier:} Enforce consistent style and reduce code review friction.
\end{itemize}

\subsection{Infrastructure and Runtime Tooling}

\begin{itemize}
    \item \textbf{Docker + Docker Compose:} Create consistent local environments and mirror production topology.
    \item \textbf{PostgreSQL:} Primary datastore for transactional integrity and query performance.
    \item \textbf{Redis:} Cache layer for AI responses and rate limiting.
    \item \textbf{Nginx:} Reverse proxy routing and request normalization.
\end{itemize}

\subsection{Development Environment Setup}

\textbf{Local environment requirements}

\begin{itemize}
    \item Node.js 18 LTS or later with npm
    \item Python 3.11 or later with pip (for the ai-service)
    \item Docker and Docker Compose v2
    \item Flutter SDK 3.x (for mobile development)
    \item PostgreSQL 15 (optional if using Docker)
    \item Git for version control
\end{itemize}

\textbf{Installation and configuration}

The development environment is configured to mirror the production topology using Docker Compose. The following steps outline the setup process:

\begin{enumerate}
    \item \textbf{Clone the repository:}
\begin{lstlisting}[language=bash, caption=none]
git clone https://github.com/yegeta/ai-learning-roadmap.git
cd ai-learning-roadmap
\end{lstlisting}
    \item \textbf{Environment variables:} Copy \texttt{.env.example} to \texttt{.env} in each service directory and configure the following:
\begin{lstlisting}[language=bash, caption=none]
# AI Service
GEMINI_API_KEY=your-gemini-api-key
GEMINI_MODEL=gemini-2.5-flash
PHI4_BASE_URL=  # Optional: URL to Phi-4 on Kaggle
OLLAMA_BASE_URL=http://host.docker.internal:11434
OLLAMA_MODEL=qwen2.5:3b

# Database
DATABASE_URL=postgresql://user:password@localhost:5432/learning_roadmap

# Authentication
JWT_SECRET=your-jwt-secret
GOOGLE_CLIENT_ID=your-google-oauth-client-id
GOOGLE_CLIENT_SECRET=your-google-oauth-client-secret
GITHUB_CLIENT_ID=your-github-oauth-client-id
GITHUB_CLIENT_SECRET=your-github-oauth-client-secret
\end{lstlisting}
    \item \textbf{Start the backend stack:}
\begin{lstlisting}[language=bash, caption=none]
docker-compose up -d
\end{lstlisting}
        This starts PostgreSQL, Redis, Nginx, and all backend services.
    \item \textbf{Run database migrations:}
\begin{lstlisting}[language=bash, caption=none]
cd backend/services/learning-service
npx prisma migrate dev
npx prisma db seed
\end{lstlisting}
    \item \textbf{Start frontend development server:}
\begin{lstlisting}[language=bash, caption=none]
cd frontend
npm install
npm run dev
\end{lstlisting}
    \item \textbf{Start mobile development (Flutter):}
\begin{lstlisting}[language=bash, caption=none]
cd flutter_mobile
flutter pub get
flutter run
\end{lstlisting}
\end{enumerate}

\textbf{Secrets management}

\begin{itemize}
    \item Environment variables are defined in \texttt{.env} files and injected via Docker Compose.
    \item OAuth keys and AI provider tokens are never stored in source code and are supplied at runtime.
    \item The \texttt{.env} files are excluded from version control via \texttt{.gitignore}.
\end{itemize}

\textbf{Version control and CI}

\begin{itemize}
    \item Git is used for source control with feature branching (main, develop, feature/*, hotfix/*).
    \item GitHub Actions automates build, lint, and test workflows (see \texttt{.github/workflows/ci.yml}).
    \item Pull requests require passing CI checks and code review before merging.
\end{itemize}

\subsection{Rationale Summary}

The chosen stack provides strong typing, consistent APIs, scalable infrastructure, and cross-platform delivery. TypeScript governs the api-gateway, learning-service, and frontend, while Python governs the ai-service --- each language applied where it has the strongest ecosystem advantage. The stack supports the project's non-functional goals by enabling test automation, reproducible Docker environments, and resilience to external AI dependencies.

\section{Developing the Solution}

This section details how the system was implemented, the coding practices that guided development, integration strategies used across services and clients, and the most important functionality with representative code excerpts.

\subsection{Coding Practices and Development Workflow}

\begin{itemize}
    \item \textbf{Layered service architecture:} Each backend service uses routes $\rightarrow$ controller $\rightarrow$ service $\rightarrow$ data access, which keeps HTTP concerns separate from business logic.
    \item \textbf{Type safety:} TypeScript types enforce contracts in the frontend, api-gateway, and learning-service. The ai-service uses Python 3.11 type annotations with Pydantic models, providing equivalent schema enforcement and editor inference in the Python codebase.
    \item \textbf{Validation at boundaries:} Requests are validated using schemas at the API boundary to avoid propagating invalid data.
    \item \textbf{Consistent error handling:} Errors are normalized using \texttt{ApiError} to produce predictable response formats and HTTP status codes.
    \item \textbf{Incremental delivery:} Features were implemented in phases: (1) authentication and RBAC, (2) ontology and domain management, (3) enrollment with learner profile capture, (4) quiz generation and gatekeeper, (5) mastery decay and spaced repetition, (6) multi-path branching, (7) SSE streaming for AI explanations and AI Instructor, (8) gamification (XP, badges, streaks), (9) learning analytics and velocity tracking, (10) completion certificates, (11) instructor and admin portals. Each phase included migrations and tests before the next began.
\end{itemize}

\subsection{Integration Strategy During Development}

\begin{itemize}
    \item \textbf{Service-to-service integration:} The learning-service calls ai-service via HTTP for both standard JSON endpoints (quiz generation) and SSE proxying (explanations, AI Instructor). Both services share Redis for caching and rate limiting.
    \item \textbf{SSE proxy pattern:} The Python FastAPI ai-service generates tokens as an async generator and returns them as a \texttt{StreamingResponse}. The Node.js learning-service proxies this SSE stream to the browser client using \texttt{ReadableStream} forwarding without buffering. Nginx SSE location blocks ensure \texttt{proxy\_buffering off} and disable gzip on stream routes, preventing accumulation at both the Python and Node.js layers.
    \item \textbf{Client-to-service integration:} The web and mobile clients authenticate through the gateway and call the learning-service with Bearer tokens. The frontend's \texttt{useAskInstructorStream} hook reads the SSE \texttt{ReadableStream} directly via the Fetch API.
    \item \textbf{Contract testing:} Responses from ai-service are schema-validated before persistence, and OAuth flows are tested for reliability.
\end{itemize}

\subsection{Major Functionality Code Excerpts}

\textbf{Authentication and OAuth flow (api-gateway)}

The gateway handles registration, login, refresh tokens, and OAuth redirection.

Source: \texttt{backend/services/api-gateway/src/modules/auth/auth.controller.ts}

\begin{lstlisting}[language=JavaScript]
export async function login(req: Request, res: Response, next: NextFunction): Promise<void> {
  try {
    const { email, password } = validate(loginSchema, req.body);
    const { user, tokens } = await authService.login(email, password);
    res.json({ user, ...tokens });
  } catch (err) {
    next(err);
  }
}

export function googleRedirect(_req: Request, res: Response): void {
  if (!config.oauth.google.clientId) {
    throw ApiError.internal('Google OAuth is not configured. Set GOOGLE_CLIENT_ID and GOOGLE_CLIENT_SECRET.');
  }
  const params = new URLSearchParams({
    client_id: config.oauth.google.clientId,
    redirect_uri: config.oauth.google.callbackUrl,
    response_type: 'code',
    scope: 'openid email profile',
  });
  res.redirect(`https://accounts.google.com/o/oauth2/v2/auth?${params.toString()}`);
}
\end{lstlisting}

\textbf{Gatekeeper logic and mastery transitions (learning-service)}

The gatekeeper enforces five-tier progression, records mastery state with the marginal-pass flag for accelerated decay, fires the correct adaptation event per failure tier, and triggers gamification and velocity recording on pass — all fire-and-forget so they never block the quiz response.

Source: \texttt{backend/services/learning-service/src/modules/gatekeeper/gatekeeper.service.ts}

\begin{lstlisting}[language=JavaScript]
export function classifyScore(scorePercent: number): GatekeeperTier {
  if (scorePercent >= 80) return 'strong_pass';
  if (scorePercent >= 70) return 'marginal_pass';
  if (scorePercent >= 50) return 'fail_low';
  if (scorePercent >= 30) return 'fail_fundamental';
  return 'fail_severe';
}

export async function applyGatekeeperOutcome(params) {
  const { userId, nodeId, enrollmentId, quizAttemptId, scorePercent } = params;
  const tier = classifyScore(scorePercent);
  const isPass = tier === 'strong_pass' || tier === 'marginal_pass';
  const isMarginalPass = tier === 'marginal_pass';
  const now = new Date();

  // Persist mastery state; isMarginalPass drives the 7-day vs 14-day decay timer
  await prisma.learnerNodeProgress.updateMany({
    where: { userId, nodeId, enrollmentId },
    data: isPass
      ? { masteryState: 'mastered', masteredAt: now, isMarginalPass, lastReviewedAt: now }
      : { masteryState: 'in_progress', lastReviewedAt: now },
  });

  // Each failure tier has its own adaptation type
  let adaptationType;
  if      (tier === 'fail_low')         adaptationType = 'resource_swap';
  else if (tier === 'fail_fundamental') adaptationType = 'prerequisite_review';
  else if (tier === 'fail_severe')      adaptationType = 'instructor_escalation';
  if (adaptationType) {
    await prisma.adaptationEvent.create({
      data: { userId, nodeId, quizAttemptId, adaptationType },
    });
  }

  if (isPass) {
    // Fire-and-forget: never block the quiz response for analytics/gamification
    const xpSource = tier === 'strong_pass'
      ? 'node_mastered_strong' : 'node_mastered_marginal';
    recordVelocity({ userId, enrollmentId, nodeId, completedAt: now }).catch(() => {});
    awardXp({ userId, source: xpSource, refId: nodeId }).catch(() => {});
    checkBadgesOnMastery({ userId, enrollmentId, nodeId, tier }).catch(() => {});
  } else {
    awardXp({ userId, source: 'quiz_attempt', refId: quizAttemptId }).catch(() => {});
  }
}
\end{lstlisting}

\textbf{Quiz generation and fallback strategy (learning-service)}

The system attempts AI generation first, then falls back to a stored quiz if needed.

Source: \texttt{backend/services/learning-service/src/modules/quizzes/quizzes.service.ts}

\begin{lstlisting}[language=JavaScript]
const freshAiQuiz = await prisma.quiz.findFirst({
  where: {
    nodeId,
    isMicroQuiz: false,
    generatedBy: 'ai_tutor',
    createdAt: { gte: new Date(Date.now() - AI_QUIZ_STALE_DAYS * 86_400_000) },
  },
  orderBy: { createdAt: 'desc' },
  include: { questions: { orderBy: { orderIndex: 'asc' } } },
});

if (freshAiQuiz) return freshAiQuiz;

const aiResponse = await requestAiQuiz({ nodeId, nodeTitle: node.title, learningOutcomes: outcomes, questionCount: 4 });
if (aiResponse?.quiz?.questions?.length) {
  return prisma.quiz.create({
    data: { nodeId, isMicroQuiz: false, generatedBy: 'ai_tutor', questions: { create: ... } },
    include: { questions: { orderBy: { orderIndex: 'asc' } } },
  });
}
\end{lstlisting}

\textbf{AI provider fallback with validation (ai-service)}

The Python AI service validates JSON outputs against a Pydantic schema and cascades across the three-tier provider chain, recording circuit-breaker outcomes at each tier.

Source: \texttt{backend/services/ai-service/app/services/ai\_service.py}

\begin{lstlisting}[language=Python]
from typing import TypeVar, Type
from app.clients.phi4_client import phi4_generate
from app.clients.ollama_client import ollama_generate
from app.clients.gemini_client import gemini_generate
from app.circuit_breaker import is_circuit_open, record_success, record_failure
from app.utils.parser import parse_and_validate

T = TypeVar("T")

async def generate(prompt: str, schema: Type[T]) -> T | None:
    # Tier 1: Phi-4-Multimodal (primary, zero-cost)
    if not await is_circuit_open("phi4"):
        raw = await phi4_generate(prompt)
        result = parse_and_validate(raw, schema)
        if result:
            await record_success("phi4")
            return result
        await record_failure("phi4")

    # Tier 2: Ollama / Qwen2.5-3B (secondary, local CPU)
    if not await is_circuit_open("ollama"):
        raw = await ollama_generate(prompt)
        result = parse_and_validate(raw, schema)
        if result:
            await record_success("ollama")
            return result
        await record_failure("ollama")

    # Tier 3: Gemini 2.5 Flash (last resort, API)
    raw = await gemini_generate(prompt)
    return parse_and_validate(raw, schema)
\end{lstlisting}

\textbf{Decay scanning and notifications (learning-service)}

The decay scanner runs on a schedule and applies Ebbinghaus-inspired thresholds: strong-pass mastery decays after 14 days, marginal-pass after 7 days (weaker initial encoding), and both decay to \texttt{relearn} at 30 days. An in-app notification is sent on each transition.

Source: \texttt{backend/services/learning-service/src/modules/decay/decay.service.ts}

\begin{lstlisting}[language=JavaScript]
const STRONG_PASS_DAYS  = 14;
const MARGINAL_PASS_DAYS = 7;
const RELEARN_DAYS       = 30;

for (const p of mastered) {
  if (!p.lastReviewedAt) continue;
  const days = (Date.now() - p.lastReviewedAt.getTime()) / 86_400_000;

  // Marginal-pass mastery uses a shorter review interval
  const threshold = p.isMarginalPass ? MARGINAL_PASS_DAYS : STRONG_PASS_DAYS;
  if (days <= threshold) continue;

  const newState = days > RELEARN_DAYS ? 'relearn' : 'review_needed';
  await prisma.learnerNodeProgress.update({
    where: { id: p.id },
    data:  { masteryState: newState },
  });
  await createNotification({
    userId: p.userId,
    type:   'decay_reminder',
    title:  `Time to review: ${p.node.title}`,
    data:   { nodeId: p.nodeId, enrollmentId: p.enrollmentId },
  });
}
\end{lstlisting}

\textbf{SSE streaming for AI explanations and AI Instructor (ai-service)}

The Python FastAPI ai-service uses async generators and \texttt{StreamingResponse} to stream tokens. Each token is emitted as an SSE \texttt{data:} event, a \texttt{[DONE]} sentinel signals completion, and \texttt{X-Accel-Buffering: no} prevents Nginx from accumulating tokens. Both the explanation and AI Instructor endpoints follow the same pattern.

Source: \texttt{backend/services/ai-service/app/routers/ai\_router.py}

\begin{lstlisting}[language=Python]
import json
from fastapi import APIRouter, Request
from fastapi.responses import StreamingResponse
from app.models.schemas import NodeContextSchema
from app.services import ai_service

router = APIRouter()

@router.post("/explain/stream")
async def stream_explanation(
    body: NodeContextSchema, request: Request
) -> StreamingResponse:
    async def event_generator():
        try:
            async for token in ai_service.stream_explanation(body):
                # Honour client disconnect
                if await request.is_disconnected():
                    break
                yield f"data: {json.dumps({'t': token})}\n\n"
        finally:
            yield "data: [DONE]\n\n"

    return StreamingResponse(
        event_generator(),
        media_type="text/event-stream",
        headers={
            "Cache-Control": "no-cache, no-transform",
            "X-Accel-Buffering": "no",   # disable Nginx buffering
            "Connection": "keep-alive",
        },
    )
\end{lstlisting}

On the frontend the React hook reads the stream token-by-token and appends each chunk to the message state, rendering the growing text in real time:

Source: \texttt{frontend/src/api/instructor-chat.ts}

\begin{lstlisting}[language=JavaScript]
const res = await fetch(`/api/v1/nodes/${nodeId}/ask/stream`, {
  method: 'POST',
  headers: { 'Content-Type': 'application/json',
             Authorization: `Bearer ${accessToken}` },
  body: JSON.stringify(payload),
  signal: controller.signal,
});
const reader = res.body.getReader();
const decoder = new TextDecoder();

while (true) {
  const { done, value } = await reader.read();
  if (done) break;
  for (const line of decoder.decode(value).split('\n')) {
    if (!line.startsWith('data: ')) continue;
    const raw = line.slice(6).trim();
    if (raw === '[DONE]') { callbacks.onDone(); return; }
    const parsed = JSON.parse(raw);
    if (parsed.t) callbacks.onChunk(parsed.t);
  }
}
\end{lstlisting}

\textbf{Certificate issuance (learning-service)}

The certificate service is idempotent (returns the existing record on double-claim), validates 100\% roadmap completion, and uses a collision-retry loop to guarantee a unique Crockford base-32 public identifier.

Source: \texttt{backend/services/learning-service/src/modules/certificates/certificates.service.ts}

\begin{lstlisting}[language=JavaScript]
function generatePublicId(): string {
  // Crockford base-32 alphabet: no I, L, O, U to avoid confusion
  const ALPHA = '0123456789ABCDEFGHJKMNPQRSTVWXYZ';
  const bytes = crypto.randomBytes(8);
  let raw = '';
  for (let i = 0; i < 8; i++) raw += ALPHA[bytes[i] % ALPHA.length];
  return `ATL-${raw.slice(0, 4)}-${raw.slice(4, 8)}`;
}

export async function claimCertificate(enrollmentId: string, userId: string) {
  // Idempotent: if already claimed, return the existing certificate
  const existing = await prisma.certificate.findUnique({ where: { enrollmentId } });
  if (existing) return serializeCertificate(existing);

  const completion = await computeCompletion(enrollmentId);
  if (completion.completionPercent < 100) {
    throw ApiError.forbidden(
      `Roadmap is ${completion.completionPercent}% complete — master every node first.`
    );
  }

  // Retry on the rare public-id collision (max 5 attempts)
  for (let attempt = 0; attempt < 5; attempt++) {
    try {
      return serializeCertificate(
        await prisma.certificate.create({
          data: {
            publicId: generatePublicId(),
            userId,
            enrollmentId,
            domainId:      enrollment.domainId,
            recipientName: enrollment.user.fullName,
            courseName:    enrollment.domain.name,
            hoursInvested: completion.completedHours,
            averageScore:  completion.averageScore,
            completedAt:   completion.lastCompletedAt ?? new Date(),
          },
        }),
      );
    } catch (e) {
      if (!isUniqueViolation(e) || attempt === 4) throw e;
    }
  }
}
\end{lstlisting}

\textbf{Gamification: XP award and level progression (learning-service)}

XP events are persisted to an append-only log for history, and a \texttt{userXp} aggregate row is upserted to keep the current total and level in sync.

Source: \texttt{backend/services/learning-service/src/modules/gamification/gamification.service.ts}

\begin{lstlisting}[language=JavaScript]
// XP weights per source (defined in gamification.types.ts)
// node_mastered_strong = 100, node_mastered_marginal = 60,
// quiz_attempt = 10, spaced_review = 25, streak_milestone = 50

export async function awardXp(params: {
  userId: string;
  source: XpSource;
  refId?: string;
}): Promise<void> {
  const amount = XP_WEIGHTS[params.source];
  if (amount <= 0) return;

  // Append-only event log for history display
  await prisma.xpEvent.create({
    data: { userId: params.userId, source: params.source,
            amount, refId: params.refId },
  });

  // Upsert the aggregate (total XP + derived level)
  const existing = await prisma.userXp.findUnique({
    where: { userId: params.userId },
  });
  const newTotal = (existing?.totalXp ?? 0) + amount;
  await prisma.userXp.upsert({
    where:  { userId: params.userId },
    create: { userId: params.userId, totalXp: newTotal,
               level: computeLevel(newTotal) },
    update: { totalXp: newTotal, level: computeLevel(newTotal) },
  });
}
\end{lstlisting}

\textbf{Roadmap visualization (web)}

The roadmap UI renders a DAG with auto-layout and color-coded mastery states.

Source: \texttt{frontend/src/features/roadmap/components/RoadmapCanvas.tsx}

\begin{lstlisting}[language=JavaScript]
const depths = useMemo(() => computeDepths(roadmapNodes, roadmapEdges), [roadmapNodes, roadmapEdges]);
const initialNodes = useMemo(() => {
  const rfNodes = toFlowNodes(roadmapNodes, depths);
  const rfEdges = toFlowEdges(roadmapEdges, nodeMap);
  return autoLayout(rfNodes, rfEdges);
}, [roadmapNodes, roadmapEdges, nodeMap, depths]);

return (
  <ReactFlow
    nodes={nodes}
    edges={edges}
    onNodeClick={handleNodeClick}
    nodeTypes={NODE_TYPES}
    fitView
  >
    <Background variant={BackgroundVariant.Dots} gap={28} size={1} color="#d6cfbf" />
  </ReactFlow>
);
\end{lstlisting}

\textbf{Token refresh and API integration (web)}

The web client injects Bearer tokens and performs silent refresh on 401 responses.

Source: \texttt{frontend/src/api/client.ts}

\begin{lstlisting}[language=JavaScript]
apiClient.interceptors.request.use((config) => {
  const token = getAccessToken?.();
  if (token) config.headers.Authorization = `Bearer ${token}`;
  return config;
});

apiClient.interceptors.response.use(
  (res) => res,
  async (error) => {
    if (error.response?.status !== 401) return Promise.reject(error);
    const { data } = await axios.post(`${apiClient.defaults.baseURL}/auth/refresh`, { refreshToken });
    original.headers.Authorization = `Bearer ${data.accessToken}`;
    return apiClient(original);
  },
);
\end{lstlisting}

\textbf{Login form validation (web)}

The login UI validates input using Zod and shows actionable error messages.

Source: \texttt{frontend/src/features/auth/components/LoginForm.tsx}

\begin{lstlisting}[language=JavaScript]
const schema = z.object({
  email: z.string().email('Invalid email address'),
  password: z.string().min(6, 'Password must be at least 6 characters'),
});
const { register, handleSubmit, formState: { errors } } = useForm<FormData>({
  resolver: zodResolver(schema),
});
\end{lstlisting}

\textbf{Mobile API client with refresh (Flutter)}

The mobile app mirrors the web token refresh pattern using Dio interceptors.

Source: \texttt{flutter\_mobile/lib/core/api/api\_client.dart}

\begin{lstlisting}[language=JavaScript, caption=none]
_dio.interceptors.add(
  InterceptorsWrapper(
    onRequest: (options, handler) {
      final accessToken = _accessTokenGetter?.call();
      if (accessToken != null && accessToken.isNotEmpty) {
        options.headers['Authorization'] = 'Bearer $accessToken';
      }
      handler.next(options);
    },
    onError: (error, handler) async {
      if (error.response?.statusCode != 401) return handler.next(error);
      final refreshedAccessToken = await _refreshAccessToken();
      if (refreshedAccessToken == null) return handler.next(error);
      final response = await _dio.request<dynamic>(requestOptions.path, options: Options(
        headers: {...requestOptions.headers, 'Authorization': 'Bearer $refreshedAccessToken'},
      ));
      handler.resolve(response);
    },
  ),
);
\end{lstlisting}

\textbf{Mobile roadmap view (Flutter)}

The mobile client renders the learning graph using GraphView and provides a node detail sheet.

Source: \texttt{flutter\_mobile/lib/features/roadmap/roadmap\_screen.dart}

\begin{lstlisting}[language=JavaScript, caption=none]
child: GraphView(
  graph: _graph,
  algorithm: _algorithm,
  paint: Paint()
    ..color = const Color(0xFFB9B1A4)
    ..strokeWidth = 1.5,
  builder: (Node node) {
    final mapped = node.key?.value as RoadmapNode;
    return RoadmapNodeWidget(
      node: mapped,
      isActive: _activeNodeId == mapped.id,
      onTap: mapped.unlocked ? () => _openNodeSheet(mapped) : null,
    );
  },
),
\end{lstlisting}

\subsection{Integration Challenges and Resolutions}

Several integration challenges emerged during development and were resolved through iterative refinement:

\begin{itemize}
    \item \textbf{AI response inconsistency:} LLM outputs varied in format. The solution was strict JSON parsing with multiple parse candidates (raw, code-block-stripped, extracted JSON slice), schema validation via Joi, and fallback providers across the three-tier chain (Phi-4 $\rightarrow$ Qwen2.5 $\rightarrow$ Gemini).
    \item \textbf{SSE buffering:} Nginx, the Node.js learning-service gzip middleware, and Python's uvicorn response buffering all independently accumulated SSE tokens before forwarding, producing batched output instead of live streaming. The fix required three coordinated changes: adding \texttt{X-Accel-Buffering: no} to FastAPI response headers, disabling gzip compression on \texttt{/stream} routes in the Node.js middleware, and configuring Nginx \texttt{proxy\_buffering off} SSE location blocks.
    \item \textbf{Performance bottlenecks:} The local Qwen2.5-3B model on CPU can take 30--90 seconds per request. Redis caching with per-type TTLs and circuit-breaker thresholds prevent repeated slow calls and isolate the performance impact.
    \item \textbf{Certificate migration conflict:} Running \texttt{prisma db push} to add the certificates table would have triggered a \texttt{UserRole} enum drift, deleting three real instructor accounts. The table was added via a targeted raw SQL migration (\texttt{add\_certificates.sql}) instead.
    \item \textbf{Authorization edge cases:} Some admin and instructor routes needed additional role checks; these were added at both gateway middleware and service layers for defence in depth.
    \item \textbf{Cross-platform consistency:} Web (React Flow) and mobile (GraphView) required separate DAG rendering implementations. Shared TypeScript API contract types ensured consistent node and edge data regardless of rendering engine.
\end{itemize}

\subsection{Summary}

The implementation preserves the core architectural design while adding pragmatic safeguards such as caching, validation, and circuit breakers. The system uses a deliberate language split: TypeScript governs the api-gateway, learning-service, and frontend (type-safe relational domain logic); Python 3.11 governs the ai-service (first-class AI ecosystem, idiomatic async generator streaming). The three-tier AI fallback strategy (Phi-4-Multimodal $\rightarrow$ Qwen2.5-3B via Ollama $\rightarrow$ Gemini 2.5 Flash) implemented in Python ensures resilience, near-zero API cost, and graceful degradation in resource-constrained environments. Beyond the originally scoped features, the implementation delivered: real-time SSE-streamed AI explanations and an AI Instructor chatbot; a verifiable completion certificate system with client-side PDF/PNG download and a public verification page; a comprehensive gamification layer (XP, level progression, streaks, achievement badges, weekly goals); learning velocity analytics with a velocity-adjusted timeline forecast; and role-stratified instructor and administrator portals. All fire-and-forget side effects (gamification, velocity recording) are dispatched asynchronously from the gatekeeper to maintain fast quiz submission response times. The system is modular, testable, and fully aligned with the functional and non-functional requirements described in Chapter~4.

\endinput
